Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-17 paper2.qmd
section:appendix; process:write

::: {#fig-example layout-ncol=2}

@@

<!-- 5. Continuing with the motivating example -->
Researchers can revisit the scenario in @sec-appendix-B1 using these insights. By applying the BNF to the SCM [-@fig-scm_example1a] and DAG [-@fig-mdag_example1a] structures, they can represent the joint probability of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$. From this expression, researchers can infer that the outcome $Y$ is not unconditionally independent of the teaching method $T$ $(Y \dsep T)$. This dependency arises from two key structures: a direct causal path from the teaching method $T$ to the outcome $Y$, represented by the two-connected-nodes structure $T \rightarrow Y$ (black path), and a confounding non-causal path from the teaching method $T$ to the outcome $Y$ through the socio-economic status of the school $X$, represented by the fork structure $T \leftarrow X \rightarrow Y$ (gray path).

These probabilistic insights also indicate that narrowing the focus to a subset of the population defined by the socio-economic status of the school $X$, achieved by conditioning on $X$ and depicted by a square in DAG [-@fig-mdag_example1a], effectively blocks all confounding non-causal path introduced by the fork structure. Nevertheless, since $Y$ remains not d-separated from $T$ even after conditioning $(Y \ndsep T \cond X)$, this lack of d-separation can only be attributed to the direct causal effect $T \rightarrow Y$.

::: {#fig-example layout-ncol=2}

::: {#fig-scm_example1a fig-align="center"}
$$
\begin{aligned}
  X & := x \\
  T & := f_{T}(x,e_{T}) \\
  Y & := f_{Y}(T,x,e_{Y}) \\
  e_{T} & \dsep e_{X} \\
  e_{T} & \dsep e_{Y} \\
  e_{X} & \dsep e_{Y}
\end{aligned}
$$

@@

![Conditioned DAG](images/png/mdag_example1a.png){#fig-mdag_example1a fig-align="center" width=65%}

@@

<!-- 4. revisiting the motivating example -->
With these new insights, researchers revisiting the scenario in @sec-appendix-B22 can infer that the socio-economic status of the school, $X$, satisfies the backdoor criterion, assuming the causal structure depicted by the SCM [-@fig-scm_example1a] and DAG [-@fig-mdag_example1a] is correct. This means that $X$ serves as a sufficient adjustment set, as it effectively blocks all confounding non-causal paths introduced by the fork structure. In this context, researchers can apply the *backdoor adjustment* to identify the ACE of $T$ on $Y$. They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE, as expressed in @eq-mACE2 (see @sec-appendix-B). Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example1a] is equivalent to the manipulated DAG [-@fig-mdag_example1b], because $X$ satisfies the backdoor criterion.

@@

::: {#fig-example layout-ncol=2}

::: {#fig-scm_example1b fig-align="center"}
$$
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := t \\
  Y & := f_{Y}(t,X,e_{Y}) \\
  e_{T} & \dsep e_{X} \\
  e_{T} & \dsep e_{Y} \\
  e_{X} & \dsep e_{Y}
\end{aligned}
$$

@@

![Manipulated DAG](images/png/mdag_example1b.png){#fig-mdag_example1b fig-align="center" width=65%}

>>>>

::: {#fig-example1 layout-ncol=2}

@@

<!-- 5. Continuing with the motivating example -->
Researchers can revisit the scenario in @sec-appendix-B1 using these insights. By applying the BNF to the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] structures, they can represent the joint probability of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$. From this expression, researchers can infer that the outcome $Y$ is not unconditionally independent of the teaching method $T$ $(Y \dsep T)$. This dependency arises from two key structures: a direct causal path from the teaching method $T$ to the outcome $Y$, represented by the two-connected-nodes structure $T \rightarrow Y$ (black path), and a confounding non-causal path from the teaching method $T$ to the outcome $Y$ through the socio-economic status of the school $X$, represented by the fork structure $T \leftarrow X \rightarrow Y$ (gray path).

These probabilistic insights also indicate that narrowing the focus to a subset of the population defined by the socio-economic status of the school $X$, achieved by conditioning on $X$ and depicted by a square in DAG [-@fig-mdag_example2], effectively blocks all confounding non-causal path introduced by the fork structure. Nevertheless, since $Y$ remains not d-separated from $T$ even after conditioning $(Y \ndsep T \cond X)$, this lack of d-separation can only be attributed to the direct causal effect $T \rightarrow Y$.

::: {#fig-example2 layout-ncol=2}

::: {#fig-scm_example2 fig-align="center"}
$$
\begin{aligned}
  X & := x \\
  T & := f_{T}(x,e_{T}) \\
  Y & := f_{Y}(T,x,e_{Y}) \\
  e_{T} & \dsep e_{X} \\
  e_{T} & \dsep e_{Y} \\
  e_{X} & \dsep e_{Y}
\end{aligned}
$$

@@

![Conditioned DAG](images/png/mdag_example1a.png){#fig-mdag_example2 fig-align="center" width=65%}

@@

<!-- 4. revisiting the motivating example -->
With these new insights, researchers revisiting the scenario in @sec-appendix-B22 can infer that the socio-economic status of the school, $X$, satisfies the backdoor criterion, assuming the causal structure depicted by the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] is correct. This means that $X$ serves as a sufficient adjustment set, as it effectively blocks all confounding non-causal paths introduced by the fork structure. In this context, researchers can apply the *backdoor adjustment* to identify the ACE of $T$ on $Y$. They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE, as expressed in @eq-mACE2 (see @sec-appendix-B). Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example2] is equivalent to the manipulated DAG [-@fig-mdag_example3], because $X$ satisfies the backdoor criterion.

@@

::: {#fig-example3 layout-ncol=2}

::: {#fig-scm_example3 fig-align="center"}
$$
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := t \\
  Y & := f_{Y}(t,X,e_{Y}) \\
  e_{T} & \dsep e_{X} \\
  e_{T} & \dsep e_{Y} \\
  e_{X} & \dsep e_{Y}
\end{aligned}
$$

@@

![Manipulated DAG](images/png/mdag_example1b.png){#fig-mdag_example3 fig-align="center" width=65%}
